Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-11-10 paper/paper.Rmd
a few more Grammarly suggested edits

library(bookdown)
library(readr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(kableExtra)
library(stringr)
```
# Introduction

@@

Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider the ways in which they can mutually reinforce each other. Even if the analysis in B&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction. Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis of ad hoc models. But this is hard to do in practice. The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc. They are useful for identifying variables that could be useful for prediction, and this is in fact already taking place.

Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning.  Even if B\&S were correct that in their study the theory-based escalation model performed better, it would simply be a case study of a single case where it was preferable. It establishes no general tendency, let alone law-like evidence for general theory. Nor, against it.

@@

1. Throughout the R code for the main analyses, the models predict a 1-time period lead version of the civil war onset variables (ending in "_plus1"). This means that, for example, in the 6-month version of the data, a value of "1" for the "_plus1" outcome variable for 2015-H2 actually refers to a civil war onset in the first half of 2016. One can confirm, through some tracing to work back to the original variable assignments, that the ROC curves and AUC are correspondingly, and correctly, also scored using the lead versions of the DV. 
2. The nominally 2016-H1 forecasts are created at the bottom of `+master.R` using the data for 2016-H1 and still the lead DV. Thus the lead DV here refers to outcomes in 2016-H2. 
3. The scoring of the forecasts, whatever time period they are actually for, is done in a separate Stata do file, `6mo_make_confusion_matrix.do`. Here, the scoring is done using the non-lead version of the DV. Thus there is a mismatch between the time period the forecasts are for and the period who's observed outcomes are used to score them.

@@

Furthermoe, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset. In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.

@@

What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated. Table \ref{tab:hp} shows the default values used by the implementation of random forest that B&S use (from the **randomForest** R package), in contrast to the basic settings used by B&S for most the models reported in the paper.

As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly. The implementation of random forests that B&S use ([@liaw:wiener:2002]) is based on the original @breiman:2001 implementation and calculates predictive probabilities by averaging over the "0" or "1" votes from all constituent decision trees. The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number. From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data. These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.

>>>>

library(bookdown)
library(readr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(kableExtra)
library(stringr)
```

# Introduction

@@

Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider how they can mutually reinforce each other. Even if the analysis in B&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction. Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis of ad hoc models. But this is hard to do in practice. The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc. They are useful for identifying variables that could be useful for prediction, and this is already taking place.

Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based on machine learning.  Even if B\&S were correct that in their study the theory-based escalation model performed better, it would simply be a case study of a single case where it was preferable. It establishes no general tendency, let alone law-like evidence for general theory. Nor, against it.

@@

1. Throughout the R code for the main analyses, the models predict a 1-time period lead version of the civil war onset variables (ending in "_plus1"). This means that, for example, in the 6-month version of the data, a value of "1" for the "_plus1" outcome variable for 2015-H2 actually refers to a civil war onset in the first half of 2016. One can confirm, through some tracing to work back to the original variable assignments, that the ROC curves and AUC are correspondingly, and correctly, also scored using the lead versions of the DV. 
2. The nominally 2016-H1 forecasts are created at the bottom of `+master.R` using the data for 2016-H1 and still the lead DV. Thus the lead DV here refers to outcomes in 2016-H2. 
3. The scoring of the forecasts, whatever time period they are actually for, is done in a separate Stata do file, `6mo_make_confusion_matrix.do`. Here, the scoring is done using the non-lead version of the DV. Thus there is a mismatch between the time period the forecasts are for and the period whose observed outcomes are used to score them.

@@

Furthermore, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset. In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.

@@

What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated. Table \ref{tab:hp} shows the default values used by the implementation of random forest that B&S use (from the **randomForest** R package), in contrast to the basic settings used by B&S for most of the models reported in the paper.

As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly. The implementation of random forests that B&S use ([@liaw:wiener:2002]) is based on the original @breiman:2001 implementation and calculates predictive probabilities by averaging over the "0" or "1" votes from all constituent decision trees. The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number. From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive and operates on a large bootstrapped sample of the original training data. These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.

### 2020-11-10 paper/paper.tex
a few more Grammarly suggested edits

Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider the ways in which they can mutually reinforce each other. Even if the analysis in B\&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction. Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis of ad hoc models. But this is hard to do in practice. The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc. They are useful for identifying variables that could be useful for prediction, and this is in fact already taking place.

Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning. Even if B\&S were correct that in their study the theory-based escalation model performed better, it would simply be a case study of a single case where it was preferable. It establishes no general tendency, let alone law-like evidence for general theory. Nor, against it.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Throughout the R code for the main analyses, the models predict a 1-time period lead version of the civil war onset variables (ending in "\_plus1``). This means that, for example, in the 6-month version of the data, a value of''1" for the "\_plus1" outcome variable for 2015-H2 actually refers to a civil war onset in the first half of 2016. One can confirm, through some tracing to work back to the original variable assignments, that the ROC curves and AUC are correspondingly, and correctly, also scored using the lead versions of the DV.
\item
  The nominally 2016-H1 forecasts are created at the bottom of \texttt{+master.R} using the data for 2016-H1 and still the lead DV. Thus the lead DV here refers to outcomes in 2016-H2.
\item
  The scoring of the forecasts, whatever time period they are actually for, is done in a separate Stata do file, \texttt{6mo\_make\_confusion\_matrix.do}. Here, the scoring is done using the non-lead version of the DV. Thus there is a mismatch between the time period the forecasts are for and the period who's observed outcomes are used to score them.
\end{enumerate}

@@

Furthermoe, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset. In the file \texttt{6mo\_make\_confusion\_matrix.do} on line 52, missing values in ``incidence\_civil\_ns'' are recoded to 1, thus reverting the onset coding of this variable to incidence.

@@

What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated. Table \ref{tab:hp} shows the default values used by the implementation of random forest that B\&S use (from the \textbf{randomForest} R package), in contrast to the basic settings used by B\&S for most the models reported in the paper.

As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly. The implementation of random forests that B\&S use ((Liaw and Wiener 2002)) is based on the original Breiman (2001) implementation and calculates predictive probabilities by averaging over the ``0'' or ``1'' votes from all constituent decision trees. The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number. From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data. These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.

>>>>

Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider how they can mutually reinforce each other. Even if the analysis in B\&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction. Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis of ad hoc models. But this is hard to do in practice. The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc. They are useful for identifying variables that could be useful for prediction, and this is already taking place.

Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based on machine learning. Even if B\&S were correct that in their study the theory-based escalation model performed better, it would simply be a case study of a single case where it was preferable. It establishes no general tendency, let alone law-like evidence for general theory. Nor, against it.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Throughout the R code for the main analyses, the models predict a 1-time period lead version of the civil war onset variables (ending in "\_plus1``). This means that, for example, in the 6-month version of the data, a value of''1" for the "\_plus1" outcome variable for 2015-H2 actually refers to a civil war onset in the first half of 2016. One can confirm, through some tracing to work back to the original variable assignments, that the ROC curves and AUC are correspondingly, and correctly, also scored using the lead versions of the DV.
\item
  The nominally 2016-H1 forecasts are created at the bottom of \texttt{+master.R} using the data for 2016-H1 and still the lead DV. Thus the lead DV here refers to outcomes in 2016-H2.
\item
  The scoring of the forecasts, whatever time period they are actually for, is done in a separate Stata do file, \texttt{6mo\_make\_confusion\_matrix.do}. Here, the scoring is done using the non-lead version of the DV. Thus there is a mismatch between the time period the forecasts are for and the period whose observed outcomes are used to score them.
\end{enumerate}

@@

Furthermore, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset. In the file \texttt{6mo\_make\_confusion\_matrix.do} on line 52, missing values in ``incidence\_civil\_ns'' are recoded to 1, thus reverting the onset coding of this variable to incidence.

@@

What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated. Table \ref{tab:hp} shows the default values used by the implementation of random forest that B\&S use (from the \textbf{randomForest} R package), in contrast to the basic settings used by B\&S for most of the models reported in the paper.

As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly. The implementation of random forests that B\&S use ((Liaw and Wiener 2002)) is based on the original Breiman (2001) implementation and calculates predictive probabilities by averaging over the ``0'' or ``1'' votes from all constituent decision trees. The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number. From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive and operates on a large bootstrapped sample of the original training data. These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.
